\section{Getting started in AI safety}
\label{sec:conclusion}

%Mathematicians do not need to become machine-learning engineers, and beautiful problems do not need safety justifications. But the profession already knows how to do something the world badly needs: turn vague pressure into definitions, invariants, examples, counterexamples, and theorems. The difficulty is choosing where that craft should go.

Technological progress is a human choice, not a law of nature. When mathematical ideas help build systems whose effects will be felt at civilizational scale, truth-seeking remains a core value, but it is no longer the only value in view. The civic mathematician also asks what our truths are for: whether they can help design AI systems to be more legible, more accountable, and easier to steer toward coexistence with us.

The open problems above are meant as entry points into the new field of Math for AI Safety. Other ways in are workshops like \href{https://www.iliadconference.com}{ILIAD}, the \href{https://www.arena.education/}{ARENA} curriculum, research programs like \href{https://www.matsprogram.org/}{MATS} and \href{https://sparai.org/}{SPAR}, and organizations like \href{https://humancompatible.ai/}{CHAI}, \href{https://constellation.org/}{Constellation}, \href{https://www.far.ai/}{FAR AI}, and \href{https://resolution.org}{Resolution}. Funding for work in this area is available from grantmakers like \href{https://coefficientgiving.org/funds/navigating-transformative-ai/}{Coefficient Giving}, \href{https://www.cooperativeai.com/}{CAIF}, \href{https://survivalandflourishing.fund/}{SFF}, and \href{https://alignmentproject.aisi.gov.uk/}{UK AISI}. % pass 2: resource list moved before the closing line so the section ends on the exhortation, not the link list
I hope you'll pick a problem whose language already feels like home, strip it down to the simplest interesting case, and start proving things!

\section*{Acknowledgments}

I thank Jesse Hoogland, Hyojeong Son, Jacob Tsimerman, Claude Fable 5, Claude Opus 4.8, and GPT 5.5 and 5.6 Sol for many inspiring conversations. Scott Aaronson, Ahmed Bou-Rabee, Vince Conitzer, Jonathan Gabor, Chris Hillar, Brad Knox, Phil Sosoe, Samuel Speas, Kate Stange, Steve Strogatz, Ariel Yadin, and an anonymous referee provided valuable feedback on an early draft.

\appendix

\section{A glossary for mathematicians}
\label{app:glossary}

\begin{longtable}{@{}p{0.26\textwidth} p{0.68\textwidth}@{}}
\glsentry{neural-network}{neural network} & a function with millions to billions of adjustable parameters (its \emph{weights}), built by composing simple layers and tuned by \emph{gradient descent} to reduce a training \emph{loss}. \\[5pt]
\glsentry{language-model}{language model} & a neural network trained to predict the next token of text; today's chat systems are language models further trained to follow instructions. \\[5pt]
\glsentry{token}{token} & the atomic unit (word or word-fragment) a language model reads and predicts. \\[5pt]
\glsentry{loss}{loss} & the scalar a network is trained to minimize; for language models, the negative log-probability it assigned to the actual next token. \\[5pt]
\glsentry{weights}{weights} & the trainable parameters of an artificial neural network: the entries of its matrices, trained by gradient descent and fixed at deployment time. \\[5pt]
\glsentry{activations}{activations} & the values that flow through the network on a given input (the outputs of its neurons and layers), as opposed to the fixed weights. \\[5pt]
\glsentry{activation-space}{activation space} & the copy of $\mathbb{R}^n$ in which a given layer's activations live, one coordinate per neuron; feature directions are directions in this space. \\[5pt]
\glsentry{gradient-descent}{gradient descent} & the optimizer: repeatedly nudge the parameters $w$ against the gradient of the loss, $w \leftarrow w - \eta\,\nabla L(w)$. \\[5pt]
\glsentry{relu}{ReLU} & the rectified linear unit $x \mapsto \max(0,x)$, a canonical piecewise-linear nonlinearity; modern transformers often use smooth or gated variants instead (GELU, SiLU, SwiGLU). \\[5pt]
\glsentry{transformer}{transformer} & the dominant network architecture for language: a stack of layers that mix information across token positions by \emph{attention} (each position selectively reads from the others) and read from / write to the residual stream. \\[5pt]
\glsentry{residual-stream}{residual stream} & the running vector of activations a transformer reads from and writes to at each layer; a natural home for feature directions. \\[5pt]
\glsentry{feature}{feature} & a hypothesized variable or property a network represents in its activations (e.g.\ ``the text is in French''). \\[5pt]
\glsentry{probe}{probe} & a simple (usually linear) function of a network's activations, trained or searched to read off some quantity of interest. \\[5pt]
\glsentry{policy}{policy} & an agent's rule for choosing actions from states (a possibly randomized map from states to actions). \\[5pt]
\glsentry{reward-hacking}{reward hacking} & raising a measured reward signal in ways that defeat the intent behind it; Goodhart's law as it appears in agents trained by reward. \\[5pt]
\glsentry{superposition}{superposition} & a hypothesized representation scheme in which a network stores more features than it has dimensions, as non-orthogonal directions, tolerable when few features are active at once. \\[5pt]
\glsentry{sparse-autoencoder}{sparse autoencoder} & a network trained to reconstruct another network's activations as sparse combinations of learned dictionary directions; an empirical tool for extracting features. \\[5pt]
\glsentry{interpretability}{interpretability} & reverse-engineering a trained network into human-understandable structure (circuits, features, algorithms), validated by intervention and not by human-readable description alone. \\[5pt]
% corrigibility row commented out 2026-07-09: its first use (the intro's control/coexistence footnote) is commented out; restore both together.
% \glsentry{corrigibility}{corrigibility} & the property that an AI system accepts correction, modification, and shutdown by its human principals, rather than resisting them. \\[5pt]
\glsentry{gradual-disempowerment}{gradual disempowerment} & the risk that humans lose influence over the institutions that shape their lives, by incremental delegation to AI systems we do not understand rather than by any sudden takeover. \\
\end{longtable}

\begin{thebibliography}{99}

% 2026-07-12: eleven uncited entries deleted outright per the citation review
% (edits/citation-recommendations.md), along with einsteinarena and sanborn2022
% (cut with their citations). Restored the same day with new one-sentence homes:
% rao1999, wald1950, wentworth2025, ngo2025scale.

\bibitem{abbe2023} Abbe, E., Boix-Adser\`a, E., and Misiakiewicz, T. (2023). SGD learning on neural networks: leap complexity and saddle-to-saddle dynamics. \emph{Conference on Learning Theory (COLT)} 2023. \arxiv{2302.11055}.

\bibitem{aharon2006} Aharon, M., Elad, M., and Bruckstein, A. M. (2006). On the uniqueness of overcomplete dictionaries, and a practical way to retrieve them. \emph{Linear Algebra and its Applications} 416(1), 48--67.

\bibitem{alon2026remarks} Alon, N., Bloom, T. F., Gowers, W. T., Litt, D., Sawin, W., Shankar, A., Tsimerman, J., Wang, V., and Wood, M. M. (2026). Remarks on the disproof of the unit distance conjecture. \arxiv{2605.20695}.

\bibitem{alquier2020} Alquier, P., and Ridgway, J. (2020). Concentration of tempered posteriors and of their variational approximations. \emph{Annals of Statistics} 48(3), 1475--1497. \arxiv{1706.09293}.

\bibitem{arditi2024} Arditi, A., Obeso, O., Syed, A., Paleka, D., Panickssery, N., Gurnee, W., and Nanda, N. (2024). Refusal in language models is mediated by a single direction. \emph{Advances in Neural Information Processing Systems} 37. \arxiv{2406.11717}.


\bibitem{barasz2014} Bar\'asz, M., Christiano, P., Fallenstein, B., Herreshoff, M., LaVictoire, P., and Yudkowsky, E. (2014). Robust cooperation in the Prisoner's Dilemma: program equilibrium via provability logic. \arxiv{1401.5577}.


\bibitem{bricken2023} Bricken, T., et al. (2023). Towards monosemanticity: decomposing language models with dictionary learning. \emph{Transformer Circuits Thread}, Anthropic. \url{https://transformer-circuits.pub/2023/monosemantic-features}.



\bibitem{caizhang2014} Cai, T. T., and Zhang, A. (2014). Sparse representation of a polytope and recovery of sparse signals and low-rank matrices. \emph{IEEE Transactions on Information Theory} 60(1), 122--132. \arxiv{1306.1154}.

\bibitem{candes2006} Cand\`es, E. J., Romberg, J., and Tao, T. (2006). Robust uncertainty principles: exact signal reconstruction from highly incomplete frequency information. \emph{IEEE Transactions on Information Theory} 52(2), 489--509.

\bibitem{candes2008} Cand\`es, E. J. (2008). The restricted isometry property and its implications for compressed sensing. \emph{Comptes Rendus Math\'ematique} 346(9--10), 589--592.




\bibitem{chughtai2023} Chughtai, B., Chan, L., and Nanda, N. (2023). A toy model of universality: reverse-engineering how networks learn group operations. \emph{International Conference on Machine Learning} 2023. \arxiv{2302.03025}.

\bibitem{cooper2025} Cooper, E., Oesterheld, C., and Conitzer, V. (2025). Characterising simulation-based program equilibria. \emph{Proceedings of AAAI} 2025. \arxiv{2412.14570}.

\bibitem{critch2016} Critch, A. (2019). A parametric, resource-bounded generalization of L\"ob's theorem, and a robust cooperation criterion for open-source game theory. \emph{The Journal of Symbolic Logic} 84(4), 1368--1381. \arxiv{1602.04184}.

\bibitem{critch2022} Critch, A., Dennis, M., and Russell, S. (2022). Cooperative and uncooperative institution designs: surprises and problems in open-source game theory. \arxiv{2208.07006}.

\bibitem{cunningham2023} Cunningham, H., Ewart, A., Riggs, L., Huben, R., and Sharkey, L. (2023). Sparse autoencoders find highly interpretable features in language models. \arxiv{2309.08600}.


\bibitem{donoho2006} Donoho, D. L. (2006). Compressed sensing. \emph{IEEE Transactions on Information Theory} 52(4), 1289--1306.

\bibitem{dyson2009} Dyson, F. (2009). Birds and frogs. \emph{Notices of the American Mathematical Society} 56(2), 212--223.



\bibitem{eisenstat2025} Eisenstat, S. (2025). Condensation: a theory of concepts. Manuscript. \url{https://www.sameisenstat.net/doc/condensation-25-07.pdf}.

\bibitem{elhage2022} Elhage, N., et al. (2022). Toy models of superposition. \emph{Transformer Circuits Thread}. \arxiv{2209.10652}.





\bibitem{garfinkle2019} Garfinkle, C. J., and Hillar, C. J. (2019). On the uniqueness and stability of dictionaries for sparse representation of noisy signals. \emph{IEEE Transactions on Signal Processing} 67(23), 5884--5892. \arxiv{1606.06997}.

\bibitem{garrabrant2016} Garrabrant, S., Benson-Tilsen, T., Critch, A., Soares, N., and Taylor, J. (2016). Logical induction. \arxiv{1609.03543}.

\bibitem{gowers2000} Gowers, W. T. (2000). The two cultures of mathematics. In \emph{Mathematics: Frontiers and Perspectives} (V. Arnold, M. Atiyah, P. Lax, and B. Mazur, eds.), American Mathematical Society, 65--78.

\bibitem{gribonval2015} Gribonval, R., Jenatton, R., and Bach, F. (2015). Sparse and spurious: dictionary learning with noise and outliers. \emph{IEEE Transactions on Information Theory} 61(11), 6298--6319. \arxiv{1407.5155}.

\bibitem{grunwald2012} Gr\"unwald, P. (2012). The safe Bayesian: learning the learning rate via the mixability gap. \emph{Algorithmic Learning Theory (ALT)} 2012, 169--183.





% 2026-07-13: entry moved here (was filed under its citation key "spherepacking2026",
% stranding Hariharan et al. between Sharkey and Tennenholtz).
\bibitem{spherepacking2026} Hariharan, S., Birkbeck, C., Lee, S., Ma, H. K. G., Mehta, B., Poiroux, A., and Viazovska, M. (2026). Progress in formalizing sphere packing in dimension~8. \arxiv{2604.23468}.

\bibitem{hillar2015} Hillar, C. J., and Sommer, F. T. (2015). When can dictionary learning uniquely recover sparse data from subsamples? \emph{IEEE Transactions on Information Theory} 61(11), 6290--6297. \arxiv{1106.3616}.

\bibitem{hironaka1964} Hironaka, H. (1964). Resolution of singularities of an algebraic variety over a field of characteristic zero. \emph{Annals of Mathematics} 79(1), 109--203; 79(2), 205--326.

\bibitem{howard1988} Howard, J. V. (1988). Cooperation in the Prisoner's Dilemma. \emph{Theory and Decision} 24(3), 203--213.



\bibitem{johnson1984} Johnson, W. B., and Lindenstrauss, J. (1984). Extensions of Lipschitz mappings into a Hilbert space. \emph{Contemporary Mathematics} 26, 189--206.

\bibitem{khan2023} Khan, M. E., and Rue, H. (2023). The Bayesian learning rule. \emph{Journal of Machine Learning Research} 24(281), 1--46. \arxiv{2107.04562}.

\bibitem{kulveit2025} Kulveit, J., Douglas, R., Ammann, N., Turan, D., Krueger, D., and Duvenaud, D. (2025). Gradual disempowerment: systemic existential risks from incremental AI development. \arxiv{2501.16946}.






\bibitem{kwa2025} Kwa, T., West, B., Becker, J., et al. (2025). Measuring AI ability to complete long software tasks. METR report; \emph{Advances in Neural Information Processing Systems} 38. \arxiv{2503.14499}.

% 2026-07-13: Orseau dropped to match the published ICML author list (he appears
% only on the arXiv version); first author kept as "Langosco" per Lionel.
\bibitem{langosco2022} Langosco, L., Koch, J., Sharkey, L., Pfau, J., and Krueger, D. (2022). Goal misgeneralization in deep reinforcement learning. \emph{International Conference on Machine Learning} 2022. \arxiv{2105.14111}.

\bibitem{lau2023} Lau, E., Furman, Z., Wang, G., Murfet, D., and Wei, S. (2025). The local learning coefficient: a singularity-aware complexity measure. \emph{Artificial Intelligence and Statistics (AISTATS)} 2025. \arxiv{2308.12108}.


\bibitem{lob1955} L\"ob, M. H. (1955). Solution of a problem of Leon Henkin. \emph{The Journal of Symbolic Logic} 20(2), 115--118.



\bibitem{marchetti2024} Marchetti, G. L., Hillar, C., Kragic, D., and Sanborn, S. (2024). Harmonics of learning: universal Fourier features emerge in invariant networks. \emph{Conference on Learning Theory (COLT)} 2024. \arxiv{2312.08550}.

\bibitem{mathlib2020} The mathlib Community. (2020). The Lean mathematical library. \emph{Certified Programs and Proofs (CPP)} 2020, 367--381. \arxiv{1910.09336}.

\bibitem{murfet2026} Murfet, D., Gordon, A., Adam, M., Wang, G., Hoogland, J., Baker, G., Snell, W., van Wingerden, S., Newgas, A., Snikkers, B., and Hitchcock, R. (2026). Spectroscopy at scale: finding interpretable structure in Pythia-1.4B. Timaeus, research report. \url{https://timaeus.co/research/2026-04-21-spectroscopy-main/}.

\bibitem{nanda2023} Nanda, N., Chan, L., Lieberum, T., Smith, J., and Steinhardt, J. (2023). Progress measures for grokking via mechanistic interpretability. \emph{International Conference on Learning Representations} 2023. \arxiv{2301.05217}.

\bibitem{ngo2025scale} Ngo, R. (2025). Towards a scale-free theory of intelligent agency. LessWrong essay. \url{https://www.lesswrong.com/posts/5tYTKX4pNpiG4vzYg/towards-a-scale-free-theory-of-intelligent-agency}.

\bibitem{oesterheld2021} Oesterheld, C., and Conitzer, V. (2021). Safe Pareto improvements for delegated game playing. \emph{International Conference on Autonomous Agents and Multiagent Systems (AAMAS)} 2021; journal version, \emph{Autonomous Agents and Multi-Agent Systems} 36 (2022).

\bibitem{openai2026unitdistance} OpenAI. (2026). An OpenAI model has disproved a central conjecture in discrete geometry. \url{https://openai.com/index/model-disproves-discrete-geometry-conjecture/}.

\bibitem{park2023} Park, K., Choe, Y. J., and Veitch, V. (2024). The linear representation hypothesis and the geometry of large language models. \emph{International Conference on Machine Learning} 2024. \arxiv{2311.03658}.

\bibitem{parr2022} Parr, T., Pezzulo, G., and Friston, K. J. (2022). \emph{Active Inference: The Free Energy Principle in Mind, Brain, and Behavior}. MIT Press.

\bibitem{pearl2009} Pearl, J. (2009). \emph{Causality: Models, Reasoning, and Inference}, 2nd ed. Cambridge University Press.

\bibitem{lehalleur2025} Pepin Lehalleur, S., Hoogland, J., Farrugia-Roberts, M., Wei, S., Gietelink Oldenziel, A., Wang, G., Carroll, L., and Murfet, D. (2025). You are what you eat: AI alignment requires understanding how data shapes structure and generalisation. \arxiv{2502.05475}.

\bibitem{rao1999} Rao, R. P. N., and Ballard, D. H. (1999). Predictive coding in the visual cortex: a functional interpretation of some extra-classical receptive-field effects. \emph{Nature Neuroscience} 2(1), 79--87.


\bibitem{richens2024} Richens, J., and Everitt, T. (2024). Robust agents learn causal world models. \emph{International Conference on Learning Representations} 2024. \arxiv{2402.10877}.


\bibitem{richens2025general} Richens, J., Everitt, T., and Abel, D. (2025). General agents need world models. \emph{International Conference on Machine Learning} 2025. \arxiv{2506.01622}.

\bibitem{saff1997} Saff, E. B., and Kuijlaars, A. B. J. (1997). Distributing many points on a sphere. \emph{The Mathematical Intelligencer} 19(1), 5--11.


\bibitem{sawin2026} Sawin, W. (2026). An explicit lower bound for the unit distance problem. \arxiv{2605.20579}.

\bibitem{saxe2014} Saxe, A. M., McClelland, J. L., and Ganguli, S. (2014). Exact solutions to the nonlinear dynamics of learning in deep linear neural networks. \emph{International Conference on Learning Representations} 2014. \arxiv{1312.6120}.

\bibitem{schelling1960} Schelling, T. C. (1960). \emph{The Strategy of Conflict}. Harvard University Press.

\bibitem{sharkey2025} Sharkey, L., et al. (2025). Open problems in mechanistic interpretability. \emph{Transactions on Machine Learning Research} 2025. \arxiv{2501.16496}.






\bibitem{tennenholtz2004} Tennenholtz, M. (2004). Program equilibrium. \emph{Games and Economic Behavior} 49(2), 363--373.


\bibitem{thurston1994} Thurston, W. P. (1994). On proof and progress in mathematics. \emph{Bulletin of the American Mathematical Society} 30(2), 161--177. \arxiv{math/9404236}.


% 2026-07-13: entry moved here (von Neumann under V; it sat between OpenAI and Park).
\bibitem{vnm1944} von Neumann, J., and Morgenstern, O. (1944). \emph{Theory of Games and Economic Behavior}. Princeton University Press.

\bibitem{wald1950} Wald, A. (1950). \emph{Statistical Decision Functions}. Wiley.

\bibitem{watanabe2009} Watanabe, S. (2009). \emph{Algebraic Geometry and Statistical Learning Theory}. Cambridge University Press.

\bibitem{watanabe2013} Watanabe, S. (2013). A widely applicable Bayesian information criterion. \emph{Journal of Machine Learning Research} 14, 867--897.

\bibitem{wentworth2025} Wentworth, J., and Lorell, D. (2025). Natural latents: latent variables stable across ontologies. \arxiv{2509.03780}.

\bibitem{zhang2017} Zhang, C., Bengio, S., Hardt, M., Recht, B., and Vinyals, O. (2017). Understanding deep learning requires rethinking generalization. \emph{International Conference on Learning Representations} 2017. \arxiv{1611.03530}.


\end{thebibliography}
