\documentclass[11pt,reqno]{amsart}

% ============================================================
%  Math for AI Safety
%  Ported to amsart (mytemplate.tex house style).
%  Citations: numbered \cite with hyperref links; bibliography is a
%  manual thebibliography. arXiv links via the \arxiv macro.
% ============================================================

\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{array}
\usepackage{longtable}
\usepackage{microtype}
\usepackage[normalem]{ulem} % \uline for glossary terms at first use; normalem keeps \emph italic
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, calc, decorations.pathreplacing, shapes.geometric}
\usepackage[colorlinks=true, linkcolor=blue!55!black, citecolor=blue!55!black, urlcolor=blue!55!black]{hyperref}
\hypersetup{pdftitle={Math for AI Safety: An Invitation for Mathematicians}, pdfauthor={Lionel Levine}}

% Wider text block (mytemplate.tex house style).
\addtolength{\oddsidemargin}{-.2in}
\addtolength{\evensidemargin}{-.2in}
\addtolength{\textwidth}{.4in}

% Clickable arXiv links (from mytemplate.tex).
\newcommand{\arxiv}[1]{\href{http://arxiv.org/abs/#1}{\texttt{arXiv:#1}}}
% Glossary terms are underlined at first use (2026-07-13, per Lionel); the intro says so.
\DeclareRobustCommand{\glsfirst}[2]{\hypertarget{first:#1}{\hyperlink{gloss:#1}{\uline{#2}}}}
\DeclareRobustCommand{\glsentry}[2]{\hypertarget{gloss:#1}{\hyperlink{first:#1}{#2}}}

\theoremstyle{plain}
\newtheorem*{theorem*}{Theorem}
\theoremstyle{definition}
% A small marker for the starred open problems.
\newcommand{\op}{\ensuremath{\bigstar}\;}
\newtheorem{openproblem}{\op Open Problem}

\title[Math for AI Safety]{Math for AI Safety: \\ An invitation for mathematicians}
\author{Lionel Levine}
\date{July 22, 2026}

\begin{document}

\begin{abstract}
Powerful AI systems raise questions that look, at first, like engineering: What should the system optimize? What has it learned? Why does it generalize? How will it behave on unfamiliar inputs? Each has a mathematical core. I organize this invitation by mathematical field, so you can turn straight to your own: logic and game theory for cooperation and bounded reasoning; probability and causality for agency and world-models; algebra and representation theory for learned features; analysis and geometry for generalization and training dynamics. New mathematics is needed to help make AI systems more legible, steerable, and cooperative. Each section ends with a concrete open problem, chosen to be small enough to start and close enough to the safety question to matter.
\end{abstract}

\maketitle

\input{sections-revised/intro}
\input{sections-revised/logic}
\input{sections-revised/probability}
\input{sections-revised/algebra}
\input{sections-revised/analysis}
%\input{sections-revised/control}       % parked: not in current compile
%\input{sections-revised/topology}      % parked: not in current compile
\input{sections-revised/conclusion}

\end{document}
